\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\geometry{a4paper, margin=1in}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{mdframed}
\usepackage{xurl}
\DeclareUnicodeCharacter{2581}{\_}



\title{Medical Discharge Summary Simplification for Patients}
\author{BTP Project Report}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}
The objective of this project is to automate the simplification of complex medical discharge summaries into plain "Lay English" that can be easily understood by patients and their families. Medical documents are often laden with technical jargon, abbreviations, and complex diagnostic codes (ICD codes) that pose a significant barrier to patient understanding. This project develops a robust pipeline to bridge this communication gap using fine-tuned Large Language Models (LLMs).

Patient comprehension of discharge instructions is a critical determinant of post-hospitalization outcomes. Studies have consistently shown that patients who understand their discharge summaries are more likely to adhere to follow-up care, take medications correctly, and recognize warning signs that require urgent attention. Despite this, discharge summaries are routinely written for clinical audiences, not patients or their families. Automating lay language conversion at scale presents a meaningful opportunity to improve healthcare equity and patient safety.

\section{Task Description}
The task involves a sequence-to-sequence transformation where the input is a technical medical discharge summary and the output is a simplified version that preserves all critical clinical information while using accessible language. Key challenges include:
\begin{itemize}
    \item \textbf{Information Preservation:} Ensuring no clinical details, medications, or surgical events are lost during simplification.
    \item \textbf{Domain Specificity:} Handling highly specialized medical terminology, including drug names, surgical procedures, and laboratory values.
    \item \textbf{Document Length:} Processing long discharge summaries that exceed the token limits of standard transformer models.
    \item \textbf{Factual Fidelity:} Avoiding hallucinations — the generation of plausible-sounding but factually incorrect medical statements — which can be dangerous in a clinical communication context.
    \item \textbf{Tone and Sensitivity:} Handling emotionally sensitive content such as end-of-life events, death declarations, and deteriorating prognoses with appropriate language while remaining clinically accurate.
    \item \textbf{Structural Coherence:} Maintaining the logical flow of a clinical narrative across multiple conditions and events without merging, reordering, or omitting distinct clinical episodes.
\end{itemize}

\section{Datasets}
\subsection{Fine-tuning Dataset: \texttt{data.tsv}}
The models are fine-tuned on the \texttt{data.tsv} dataset, which contains parallel pairs of original medical texts and their simplified English counterparts. To ensure the model learns simplification rather than aggressive summarization, a preservation filter is applied to discard samples where the simplified text is significantly shorter than the original (ratio $< 0.45$). Due to computational resource constraints on the Kaggle platform—specifically GPU memory and session time limits—fine-tuning was restricted to a subset of 10,000 samples.

\subsection{Inference Dataset: \texttt{anotated\_dataset\_v2.xlsx}}
The final evaluation is performed on the \texttt{anotated\_dataset\_v2.xlsx} dataset. This contains real-world discharge summaries along with patient outcomes and ICD-10 codes. The pipeline expands abbreviations and applies a medical lay-dictionary before passing the text to the fine-tuned LLM.

\section{Methodology}
The pipeline consists of several stages:
\begin{enumerate}
    \item \textbf{Preprocessing:} Normalization and expansion of abbreviations.
    \item \textbf{Lay-Dictionary Mapping:} Replacing known medical terms with simplified Indian English equivalents (e.g., ``hypertension'' $\rightarrow$ ``high BP'').
    \item \textbf{Sliding Window Chunking:} Since most models have a 512 or 1024 token limit, long summaries are split into overlapping chunks (450 tokens with 50-token overlap) to ensure continuity and prevent information loss at boundaries.
    \item \textbf{Chunked Generation:} Each chunk is processed independently by the model, and the results are stitched together to form the final document.
\end{enumerate}

\section{Models Explored}

\subsection{BioGPT}
\textbf{Architecture:} BioGPT follows the Transformer decoder-only backbone, specifically adopting the GPT-2 medium configuration with 24 layers, a hidden size of 1024, and 16 attention heads. The final model contains approximately 347 million parameters.

\textbf{Pre-training Data:} Unlike models that undergo continuous pre-training, BioGPT was pre-trained from scratch on a massive dataset of 15 million PubMed abstracts (including both titles and abstracts). A specialized in-domain vocabulary was constructed using Byte Pair Encoding (BPE) on this corpus, resulting in a vocabulary size of 42,384, which is highly optimized for biomedical terminology.

\textbf{Why BioGPT for this task:}
\begin{itemize}
    \item \textbf{Generative Excellence:} As a causal LM, it is highly effective at generating fluent, natural language.
    \item \textbf{From-Scratch Domain Expertise:} Pre-training from scratch on PubMed allows the model to learn biomedical semantics without the interference of general-domain biases.
    \item \textbf{State-of-the-Art Performance:} As demonstrated in Luo et al. (2022), BioGPT achieves superior performance on various biomedical generation and mining tasks compared to general-purpose GPT models.
\end{itemize}
\textbf{Fine-tuned Checkpoint:} \url{https://huggingface.co/11Raghav/BioGPT}

\subsection{SciFive}
\textbf{Architecture:} SciFive is a domain-specific model based on the T5 (Text-to-Text Transfer Transformer) encoder-decoder architecture. It was initialized from the pre-trained weights of the base T5 model and then specifically adapted for the biomedical domain.

\textbf{Pre-training Data:} As detailed in Phan et al. (2021), SciFive underwent extensive re-training for an additional 200,000 steps (for the base version) on a combination of the general-domain C4 corpus, PubMed abstracts, and the PubMed Central (PMC) full-text corpus. This multi-corpus approach ensures the model retains general linguistic knowledge while mastering dense clinical and technical narratives.

\textbf{Why SciFive for this task:} 
\begin{itemize}
    \item \textbf{Text-to-Text Versatility:} By treating simplification as a text-to-text problem (similar to translation), SciFive is inherently more flexible than encoder-only models like BERT.
    \item \textbf{Complex Output Handling:} SciFive has demonstrated superior performance on biomedical tasks requiring long, coherent outputs (e.g., BioASQ question answering), making it well-suited for the detailed rewriting required in discharge summary simplification.
    \item \textbf{PMC Full-Text Advantage:} Pre-training on full PMC articles, rather than just abstracts, provides the model with better context for the narrative flow found in medical reports.
\end{itemize}
\textbf{Fine-tuned Checkpoint:} \url{https://huggingface.co/11Raghav/SciFive}

\subsection{BioBART}
\textbf{Architecture:} BioBART is based on the BART (Bidirectional and Auto-Regressive Transformers) architecture. It utilizes a bidirectional encoder and an auto-regressive decoder, making it highly effective for sequence-to-sequence tasks.

\textbf{Pre-training Data:} As detailed in Yuan et al. (2022), BioBART is initialized from the general-purpose BART-base/large checkpoints and continuously pre-trained on the PubMed abstracts corpus (approx. 41 GB of text). A key technical finding in the paper is that using only the \textit{text-infilling} task for adaptation yields better performance on biomedical NLG tasks than including sentence permutation.

\textbf{Why BioBART for this task:} 
\begin{itemize}
    \item \textbf{Seq2Seq Mastery:} The encoder-decoder nature of BART is specifically designed for rewriting and summarization, which maps directly to the simplification goal.
    \item \textbf{Domain Adaptation:} Its continuous pre-training on biomedical corpora allows it to maintain medical entity integrity while generating fluent English.
\end{itemize}
\textbf{Fine-tuned Checkpoint:} \url{https://huggingface.co/11Raghav/BioBART}

\subsection{Flan-T5}
\textbf{Architecture:} Flan-T5 is an instruction-tuned version of the T5 (Text-to-Text Transfer Transformer) encoder-decoder architecture. The version used in this project, Flan-T5-base, contains approximately 250 million parameters.

\textbf{Pre-training Data:} As described by Chung et al. (2022), Flan-T5 was fine-tuned on the FLAN (Finetuned Language Net) collection, which was scaled to 1,836 tasks across 473 datasets and 146 task categories. A critical component of its training was the inclusion of \textit{chain-of-thought} (CoT) data, which significantly improves the model's reasoning and instruction-following performance on unseen tasks.

\textbf{Why Flan-T5 for this task:} 
\begin{itemize}
    \item \textbf{Instruction Scaling:} The massive variety of instruction-based tasks in its pre-training allows Flan-T5 to generalize exceptionally well to specific simplification prompts.
    \item \textbf{Robustness:} It provides strong zero-shot and few-shot performance, often outperforming much larger models that lack instruction tuning.
    \item \textbf{Reliability:} Compared to ClinicalT5, Flan-T5 demonstrated superior stability in following the "preserve every detail" constraint without degenerating into repetitive output.
\end{itemize}
\textbf{Fine-tuned Checkpoint:} \url{https://huggingface.co/11Raghav/FlanT5}

\subsection{Pegasus-PubMed}
\textbf{Challenges:} Pegasus-PubMed was explored due to its specialized architecture for abstractive summarization. However, it was found to be unsuitable for the Kaggle environment in this project.

\textbf{Architecture \& Pre-training:} As detailed in Zhang et al. (2020), Pegasus utilizes a standard Transformer encoder-decoder architecture with approximately 568 million parameters (in the large version). Its defining feature is the \textit{Gap Sentences Generation} (GSG) pre-training objective, where entire important sentences are masked from a document and the model is tasked with generating them. This objective is designed to narrow the gap between pre-training and the downstream task of abstractive summarization.

\textbf{Why it failed in this context:} 
\begin{itemize}
    \item \textbf{Memory Intensity:} Due to its large parameter count and the complexity of the GSG-informed attention mechanism, the model consistently triggered \texttt{OutOfMemoryError} (OOM) on the Kaggle T4 x2 GPU environment. \sloppy
    \item \textbf{Input Scaling:} The model's efficiency on long clinical documents was limited by the high VRAM requirements of its specific Transformer implementation when processing documents at the 512-1024 token scale.
\end{itemize}

\subsection{ClinicalT5}
\textbf{Challenges:} ClinicalT5 was initially a primary candidate given its specialization in clinical notes. However, it was ultimately discarded due to technical failures during inference.
\begin{itemize}
    \item \textbf{Output Degeneration:} The model suffered from a severe ``boundary bug'' during chunked generation. When input token sequences were stitched together, the SentencePiece tokenizer's word-boundary markers (\texttt{▁}) were corrupted, causing the model to degenerate and output repeating characters (e.g., ``sssss'').
    \item \textbf{Instruction Fragility:} Compared to instruction-tuned models like Flan-T5, ClinicalT5 was less robust at following the complex simplification prompts, often resulting in outputs that were inaccurate.
\end{itemize}


\subsection{Cloud-based Large Language Models (LLMs)}
To overcome the factual hallucination and reasoning limitations observed in the fine-tuned biomedical models, we integrated state-of-the-art cloud-based LLMs using leading inference APIs: Groq (hosting Llama models), Cerebras, and Google GenAI (hosting Gemini 2.5 Flash).

\textbf{Llama 3 Sequence Models (via Groq and Cerebras):} We utilized Llama 3.1 8B and Llama 3.3 70B via the Groq and Cerebras inference engines. These models provide ultra-fast inference and strong instruction-following capabilities. Their immense context windows easily accommodated entire medical summaries without the need for sliding-window chunking.

\textbf{Gemini 2.5 Flash:} Google's Gemini 2.5 Flash was utilized due to its massive context window and strong zero-shot reasoning capabilities. Gemini proved highly effective at generating concise, factual lay-language translations without truncating lengthy clinical descriptions.

\section{Prompting Strategies}
To optimize the outputs of the cloud LLMs, we engineered a robust system prompt and employed varying in-context learning strategies.

\subsection{System Prompt Design}
We iteratively developed a highly constrained 12-rule system prompt to enforce accuracy, tone, and conciseness, while minimizing hallucinations. 

\begin{mdframed}[backgroundcolor=gray!10, linewidth=0.5pt]
\small
\textbf{System Prompt:}\\
You are a medical language simplification assistant. Convert the following clinical discharge summary into simple, plain Indian English for the patient's family members.\\

Rules:\\
1. Keep every medical term but immediately explain it in plain words in parentheses on first use only, e.g. 'hypertension (high BP)'.\\
2. Use simple language a 6th-grader can understand. Avoid jargon. Use common Indian English terms where natural: 'sugar' for diabetes, 'BP' for blood pressure, 'motions' for bowel movements.\\
3. Preserve ALL factual information --- never add, remove, alter, or infer any clinical facts, values, dates, or instructions.\\
4. Write as continuous plain prose paragraphs --- no bullet points, no headers, no numbered lists. Output only the simplified text, nothing else.\\
5. For any measurement (e.g. BP, blood sugar, heart rate), briefly state what the normal range is and whether the patient's value was within it.\\
6. Use a calm, respectful tone. Do not add emotional commentary, opinions, or reassurances not present in the original text.\\
7. CRITICAL --- person and tense: First check whether the summary records the patient's death. If yes, write in third person past tense for the family. If no, address the patient directly in second person.\\
8. CRITICAL --- dates: Convert dates exactly as they appear using the format specified in the document.\\
9. For each medication, state its name and purpose in plain language.\\
10. For each procedure or device, briefly state what it is and why it was done.\\
11. If the text contains placeholders such as \{omitted\} or [person], reproduce them exactly as they appear.\\
12. Be highly concise. Deliver a direct, brief summary without getting bogged down in excessive medical minutiae.
\end{mdframed}

\subsection{In-Context Learning Approaches}
To further align the model with our expectations, three different prompting methods were implemented:
\begin{itemize}
    \item \textbf{Zero-Shot:} The model relies purely on the system prompt and instructions without any prior examples. Driven by our strict 12-rule system prompt, zero-shot performance proved exceptionally reliable on powerful models like Gemini 2.5 Flash and Llama 3.3 70B, effectively maintaining factual constraints automatically.
    \item \textbf{One-Shot:} A single high-quality example of a technical discharge summary and its ideal lay-language translation is appended to the prompt. This helps anchor the model's structural expected output (e.g., continuous prose without bullet points).
    \item \textbf{Few-Shot:} Multiple curated examples are included to demonstrate varied clinical scenarios (e.g., routine recovery vs. critical deterioration). This method provided the highest consistency in formatting and tone across diverse and edge-case medical cases, teaching the model implicitly how to handle structural variations in the input data.
\end{itemize}

\section{Web Application and User Interface}
To make the medical simplification models accessible to end-users (such as hospital staff, patients, and their families), we developed a full-stack web application. The application provides an intuitive interface for inputting clinical texts and retrieving simplified explanations without restricting the user to a terminal or code environment.

\subsection{System Architecture}
The system follows a modern client-server architecture:
\begin{itemize}
    \item \textbf{Backend:} A RESTful API server built using Python and Flask. It acts as a central router, handling secure authentication, text preprocessing, sliding-window chunking logic for local models, and dynamic API request routing to various inference engines (Local PyTorch, Groq, Cerebras, and Google GenAI).
    \item \textbf{Frontend:} A single-page application (SPA) built using React, Vite, and Tailwind CSS. The interface is highly responsive, lightweight, and designed to abstract away the underlying technical complexities.
\end{itemize}

\subsection{Key Features}
The web application incorporates several features to enhance usability, security, and flexibility:
\begin{itemize}
    \item \textbf{Model \& Strategy Selection:} Users can dynamically choose their preferred AI execution engine (e.g., Gemini 2.5 Flash, Llama 3.3 via Groq, or customized local models) and prompting strategy (Zero-shot, One-shot, or Few-shot) through an intuitive dashboard dropdown.
    \item \textbf{Secure Authentication:} The platform includes a secure login and registration system governed by JSON Web Tokens (JWT). This ensures that sessions are authenticated and access to the simplification tool is securely managed.
    \item \textbf{Downloadable PDF Reports:} Once a medical summary is simplified, users can instantly export the resulting plain Indian English explanation as a formatted PDF document, making it easy to print, share, or store for personal reference.
    \item \textbf{Real-time Processing Feedback:} To accommodate the processing time required for LLM inference (particularly hardware-intensive local chunking), the interface features interactive loading states to provide continuous visual feedback while the models generate responses.
\end{itemize}

\section{Quantitative Evaluation: Readability and Compression}

To objectively measure the performance of the four fine-tuned models, we conducted a quantitative analysis on a test set of 200 medical discharge summaries. The evaluation focused on two key dimensions: \textbf{Readability}, measured using the Flesch Reading Ease score, and \textbf{Compression}, measured by the change in word count and the compression ratio.

The Flesch Reading Ease score indicates how easy a text is to understand; higher scores indicate material that is easier to read. For context, a score of 60-70 is considered standard English, while scores below 30 are typical of highly technical or academic documents. The original medical summaries in our test set had an average Flesch score of \textbf{20.24}, reflecting their high technical complexity.

Table~\ref{tab:quantitative} summarizes the results across all models.

\begin{table}[h]
\centering
\caption{Quantitative comparison of model performance on 200 summaries.}
\label{tab:quantitative}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Avg. Flesch} & \textbf{Flesch Improv.} & \textbf{Avg. Words} & \textbf{Comp. Ratio} \\
\midrule
Original Text (Baseline) & 20.24 & --- & 598.7 & 1.000 \\
\midrule
\textbf{BioGPT} & \textbf{61.77} & \textbf{+41.53} & \textbf{567.9} & \textbf{0.946} \\
\textbf{SciFive} & 53.61 & +33.37 & 668.2 & 1.108 \\
\textbf{BioBART} & 47.32 & +27.08 & 656.1 & 1.086 \\
\textbf{Flan-T5} & 12.69 & -7.55 & 781.7 & 1.327 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Analysis of Quantitative Results}
\begin{itemize}
    \item \textbf{BioGPT as the Top Performer:} BioGPT achieved the highest average Flesch score (61.77) and the greatest improvement (+41.53). It was the only model that consistently reduced the word count (compression ratio of 0.946), suggesting a more efficient simplification process that avoids redundant technical detail.
    \item \textbf{SciFive and BioBART:} Both SciFive and BioBART showed significant improvements in readability. However, they both tended to \textit{expand} the text (ratios $> 1.0$), likely due to their tendency to add explanatory phrases or maintain more of the original clinical structure.
    \item \textbf{Flan-T5 Failure:} Flan-T5 performed poorly in this quantitative evaluation, actually reducing the average Flesch score to 12.69. This is consistent with our qualitative observation that Flan-T5 often copied the original text verbatim or introduced garbled content, both of which negatively impact readability metrics.
    \item \textbf{Universality of Improvement:} While BioGPT, SciFive, and BioBART improved readability in nearly all cases, the qualitative failures (hallucinations) noted in Section~\ref{sec:qualitative} remain a critical concern that these automated metrics do not capture.
\end{itemize}

\section{Qualitative Evaluation on a Representative Example}
\label{sec:qualitative}

\subsection{Input Discharge Summary}

To qualitatively evaluate and compare all models, we present their outputs on a representative discharge summary involving a complex, multi-morbidity case with a fatal outcome. This example is particularly informative because it tests the model's ability to handle: multiple simultaneous diagnoses, escalating clinical deterioration, critical interventions (resuscitation, cardioversion, dialysis), and a death declaration — all of which require precise, sensitive, and factually accurate lay language conversion.

\begin{mdframed}[backgroundcolor=gray!10, linewidth=0.5pt]
\small
\textbf{Original Discharge Summary:}\\
admission: 12/06, discharge: 12/21, stay: 14 d 11 h, location: ward, payer: sus. 1. heart failure profile c dobutamine on 12/07 and 12/08, 2. abscess in the left thigh, osteomyelitis in the pubis, fournier's syndrome, drainage on 12/03 with ultrasound and drain placement on 12/15, treatment with ceftriaxone guided by culture of abscess escherichia coli multi s colonized antibiotic therapy after shock, incision in intensive care unit on 12/17 with abundant purulent secretion, \{omitted\}. 3. septic shock from cutaneous urinary focus on 12/16 polymyxin d0 12/15 amikacin and vancomycin on 12/17, 4. chronic kidney disease acute dialysis by fistula on 12/16 5. cardiorespiratory arrest with pulseless electrical activity, 10 minutes after intubation report of previous bradycardia, follow-up appointment after 1 minute of cardiopulmonary resuscitation, 6. vt with pulse after cardiorespiratory arrest, effective electrical cardioversion with a shock. [person] presenting with instability in the last half hour, requiring norepinephrine elevation up to 1.5 mcg/kg/min. progresses with bradycardia, qrs widening, and worsening shock. given the hyperkalemia even after hemodialysis, 2 ampoules of calcium gluconate and 100 ml of bicarbonate were administered, and the respiratory rate was increased on the ventilator in order to reduce pco2 and consequently k. 2 ampoules of atropine were administered without benefit. [person] progressed to asystolic cardiorespiratory arrest, \{omitted\} in refractory shock. i declare \{omitted\} dead, due to absence of central pulses, pupils in fixed mydriasis, absence of heart sounds and asytole on the monitor, as well as absence of brainstem reflexes. the body was released. i request contact with the family. i await the death certificate to be completed. death. drainage of fournier's gangrene. death.
\end{mdframed}

\medskip
\noindent The following subsections present each model's output, followed by a detailed error analysis.

\subsection{Reference: Expected Lay Language Summary}

For reference, the key clinical facts that a correct lay language conversion must preserve and accurately communicate are:

\begin{enumerate}
    \item Admission on December 6, discharge (death) on December 21, 14-day stay.
    \item Heart failure treated with dobutamine (a medication to support heart pumping) on December 7 and 8.
    \item A pus-filled abscess in the left thigh, bone infection (osteomyelitis) in the pubic bone, and Fournier's gangrene (a severe, life-threatening infection of the genital/perineal area). Drainage performed on December 3 via ultrasound, with a drain placed on December 15. Antibiotic ceftriaxone was used based on bacterial culture results identifying \textit{Escherichia coli}. A surgical incision was performed in the ICU on December 17.
    \item Septic shock (life-threatening infection spreading to the bloodstream) originating from a skin/urinary source on December 16. Treated with polymyxin (from December 15), amikacin, and vancomycin (from December 17).
    \item Chronic kidney disease requiring emergency dialysis via a fistula on December 16.
    \item Cardiac arrest (heart stopped) with no electrical pulse detectable, occurring 10 minutes after a breathing tube was inserted. The patient was resuscitated with CPR (chest compressions) for 1 minute.
    \item Ventricular tachycardia (a dangerous fast heart rhythm) after resuscitation, successfully treated with an electrical shock (cardioversion).
    \item Final deterioration: dangerously low blood pressure requiring high-dose norepinephrine (a medication to raise blood pressure), abnormal heart rhythm on the monitor (QRS widening), and severely elevated potassium in the blood (hyperkalemia) resistant to dialysis. Calcium gluconate (2 vials) and bicarbonate (100 ml) were given to protect the heart; ventilator breathing rate was increased to lower potassium indirectly. Atropine (2 vials) was given to treat slow heart rate without effect.
    \item The patient progressed to a final, irreversible cardiac arrest (asystole). Death was declared based on absence of central pulses, fixed dilated pupils (mydriasis), absence of heart sounds, flatline on monitor, and absence of brainstem reflexes. The family was contacted and the death certificate was completed.
\end{enumerate}

\subsection{BioBART Output and Error Analysis}

\begin{mdframed}[backgroundcolor=red!5, linewidth=0.5pt]
\small
\textbf{BioBART Output:}\\
This report describes a patient who was admitted to the hospital on 12/06. She had several medical problems, including heart failure, osteomyelitis, and fournier's syndrome. She was treated with ceftriaxone guided by a culture of the swelling, antibiotics for septic shock, and incision in the intensive care unit (ICU). She also had septic shock from a focus in the urinary system on 12/07 and 12/08. She also had chronic kidney disease (chronic kidney disease) on 12/15. She had a sudden drop in her heart rate 10 minutes after being intubated, and she was able to return to her doctor after 1 minute of cardiopulmonary regovernment health schemecitation. The report also showed that the patient was being treated with calcium gluconate and bicarbonate, and the respiratory rate was increased to reduce pco2 and heart rate. She was also treated with atropine, a muscarinic receptor antagonist. The report concludes that this patient is a very rare case of death, and that it's important to carefully consider the patient's wishes when treating this serious condition. The patient was released from the hospital and contacted with the medical team. This case is unusual because the body was released from the hospital. The patient was able to communicate with the family, but the death certificate had to be completed. The death was caused by fournier's gangrene, a type of skin disease that affects the skin.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Incorrect gender assignment:} The original summary contains no gender information, yet BioBART incorrectly refers to the patient as ``she'' throughout. This is a hallucination that introduces inaccurate personal information.
    \item \textbf{Wrong dates for septic shock:} The septic shock occurred on 12/16, but BioBART incorrectly states it occurred on ``12/07 and 12/08'' — dates that correspond to the dobutamine treatment for heart failure. This is a critical factual error caused by cross-contamination of information across clinical events.
    \item \textbf{Wrong date for kidney disease:} The acute dialysis for chronic kidney disease occurred on 12/16, but the model states ``12/15,'' which was actually the date of drain placement for the abscess. Again, dates are confused across clinical events.
    \item \textbf{Abscess omitted:} The pus-filled abscess in the left thigh — a primary source of infection — is not mentioned in the output. This is a significant omission of a clinically important finding.
    \item \textbf{Garbled text (``regovernment health schemecitation''):} The phrase describing cardiopulmonary resuscitation (CPR) is rendered as meaningless text. This reflects a tokenization or decoding error, likely occurring at a chunk boundary, producing output that is not only unintelligible but potentially alarming to a patient's family.
    \item \textbf{Incorrect explanation of respiratory rate increase:} The model states breathing rate was increased ``to reduce pco2 and heart rate.'' The actual clinical goal was to reduce pCO2 in order to lower elevated blood potassium (hyperkalemia) — not to reduce heart rate. This is a clinically dangerous misinterpretation.
    \item \textbf{Atropine described as a ``muscarinic receptor antagonist'':} Instead of simplifying the term, BioBART introduces a more technical pharmacological descriptor. This is the opposite of the intended lay language conversion and is unsuitable for a patient-facing document.
    \item \textbf{Death trivialized and misrepresented:} The model describes the death as ``a very rare case'' and states ``the patient was released from the hospital and contacted with the medical team'' — implying the patient was alive and discharged. This is a severe misrepresentation of the clinical outcome. Additionally, characterizing Fournier's gangrene as merely ``a type of skin disease that affects the skin'' grossly understates the severity of a life-threatening necrotizing infection.
    \item \textbf{Missing critical interventions:} Dobutamine infusion, polymyxin, amikacin, vancomycin, calcium gluconate dose, bicarbonate dose, electrical cardioversion, and the CPR timeline are either omitted or inaccurately described.
\end{itemize}

\textbf{Overall Assessment:} BioBART's output is factually unreliable, contains hallucinated content, confuses dates across clinical events, and catastrophically misrepresents the patient's death. It is not suitable for lay language conversion of discharge summaries in its current fine-tuned state.

\subsection{BioGPT Output and Error Analysis}

\begin{mdframed}[backgroundcolor=red!5, linewidth=0.5pt]
\small
\textbf{BioGPT Output:}\\
This text is about a patient who was admitted to the hospital. She was sent home from the hospital and stayed in the hospital for an average of 14 days. She was diagnosed with heart failure, a condition where the heart pump doesn't work properly. She also had a swelling in her left thigh, bone infection, and a condition called Fournier's syndrome. She was treated with a medicine called ceftriaxone, which is a type of antibiotic. She also received a medicine called polymyxin d0, a type of medication that helps kill bacteria. She was given a medicine called vancomycin, which is used to treat infections in the urinary tract. She was also treated with a diuretic called calcium gluconate and bicarbonate. She recovered from the infection and was able to breathe normally. However, she also had problems with her heart rate, blood pressure, and breathing. This case highlights the importance of being aware of potential problems when treating patients, especially when they are in a critical condition. It also highlights the need for close communication between the patient and their family to ensure they receive the best possible care. The authors recommend that the patient's family be involved in the decision-making process to ensure the best possible treatment. This is a case report of a patient who died from a condition called ``driasis,'' which is a condition where the body doesn't have enough blood flow. The patient didn't have any heart sounds or any movement on their monitor, and they didn't show any signs of movement in the brain. This is a rare case of a patient dying from [truncated]
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Incorrect gender assignment:} As with BioBART, BioGPT incorrectly assigns a female gender (``she'') to a patient whose gender is not specified in the source text — an unprompted hallucination.
    \item \textbf{Contradictory statements about outcome:} The model first states ``she recovered from the infection and was able to breathe normally,'' then later states the patient died. These two statements are directly contradictory within the same output. A patient's family reading this would be confused and misled.
    \item \textbf{Hallucinated condition ``driasis'':} The model invents a fictional medical condition called ``driasis,'' described as ``a condition where the body doesn't have enough blood flow.'' This is a hallucination — the original text refers to ``fixed mydriasis'' (fixed dilation of the pupils, a sign of brain death), which is a clinical finding, not a cause of death.
    \item \textbf{Incorrect description of calcium gluconate:} The model refers to calcium gluconate as ``a diuretic'' (a medication that increases urine output). Calcium gluconate is actually a cardiac-protective agent used to counteract the effects of hyperkalemia on the heart. This is a serious factual error that could cause misunderstanding of the patient's treatment.
    \item \textbf{Incorrect description of vancomycin:} The model states vancomycin is ``used to treat infections in the urinary tract.'' While vancomycin is a broad-spectrum antibiotic, in this case it was used for the treatment of septic shock and bacterial infection — not specifically a urinary tract infection. The simplification introduces a misleading specificity.
    \item \textbf{Output truncated mid-sentence:} The output ends abruptly with ``This is a rare case of a patient dying from'' — the generation was cut off, likely due to a token limit being reached. This means the model failed to complete the simplification of the full document.
    \item \textbf{Discharge summary written as a ``case report'':} BioGPT reframes the output as an academic case report (``The authors recommend...''), adopting a writing style completely inappropriate for a patient-facing document. This is a domain confusion failure, likely arising from its pre-training on PubMed literature.
    \item \textbf{Missing critical clinical details:} The abscess drainage procedures, the drain placement date, the ICU incision, amikacin, norepinephrine dosing, CPR details, QRS widening, hyperkalemia management, atropine, bicarbonate, cardioversion, and the death declaration criteria are all absent from the output.
    \item \textbf{Inappropriate moralizing:} The output inserts unsolicited commentary about ``the importance of being aware of potential problems'' and family involvement in decision-making. These additions are not present in the source text and are inappropriate for a factual discharge summary conversion.
\end{itemize}

\textbf{Overall Assessment:} BioGPT conflates the task of lay language simplification with that of writing a biomedical case report, which is a direct consequence of its exclusive pre-training on PubMed abstracts. It hallucinates conditions, contradicts itself on patient outcome, and truncates its output. It is not suitable for this task in its current state.

\subsection{Flan-T5 Output and Error Analysis}

\begin{mdframed}[backgroundcolor=red!5, linewidth=0.5pt]
\small
\textbf{Flan-T5 Output:}\\
admission: 12/06, sent home from hospital: 12/21, stay: 14 d 11 h, location: ward, payer: sus. 1. heart failure profile c dobutamine on 12/07 and 12/08, 2. pus-filled swelling in the left thigh, osteomyelitis in the pubis, fournier's syndrome, drainage on 12/03 with ultrasound and drain placement on 12/15, treatment with ceftriaxone guided by culture of pus-filled swelling escherichia coli multi s colonized infection-fighting medicine therapy after shock, incision in ICU (serious care room) on 12/17 with abundant purulent secretion, omitted 4. chronic kidney disease acute dialysis by fistula on 12/16 5. cardiorespiratory arrest with pulseless electrical activity, 10 minutes after intubation report of previous slow heartbeat, return visit to doctor appointment after 1 minute of cardiopulmonary resuscitation, 6. vt with pulse after cardioreaspirary arrest, effective electrical cardioversion with a shock. [...] Given the Hyperkalemia even following kidney diazepam, 2 ampuels of calcium Gluconate were administered [...] 3 ampoules were given without benefit. [person] progressed with asysTOLIC cardiorespritory arrest, Omitted in Refractory Shock. I Declare Omitted Dead, Due to Abstinence. [...] Simplify the following medical discharge summary into plain English that a patient's family can understand [...] Death. Drainage of fournier's gangrene. Death. , . , . , . . ' , '. ', '. '' ; : . ; . : .. [truncated]
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Near-verbatim copying of source text:} Flan-T5 largely reproduces the original discharge summary with only superficial substitutions (e.g., ``abscess'' $\rightarrow$ ``pus-filled swelling''). This is not lay language conversion — it is near-copy output. The structural complexity, clinical notation format, and numbered list style of the original are preserved entirely, failing the core task objective.
    \item \textbf{Hallucinated drug ``kidney diazepam'':} The model generates the phrase ``kidney diazepam'' as a substitution for hemodialysis (kidney dialysis). Diazepam is a sedative medication with no clinical relevance to dialysis or potassium management. This is a dangerous hallucination that introduces a completely incorrect medication name into a clinical record.
    \item \textbf{Incorrect ampoule count:} The original text states 2 ampoules of atropine were administered. Flan-T5 changes this to ``3 ampoules,'' an incorrect factual alteration of a medication dosage, which is clinically significant.
    \item \textbf{Death cause changed to ``Abstinence'':} The death declaration states the cause as absence of central pulses, fixed pupils (mydriasis), and absence of brainstem reflexes. Flan-T5 replaces this entirely with ``Due to Abstinence'' — a completely fabricated and medically nonsensical cause of death. This is perhaps the most dangerous error in any of the model outputs, as it directly falsifies the documented cause of death.
    \item \textbf{Prompt leakage into output:} The model reproduces its own instruction prompt (``Simplify the following medical discharge summary into plain English...'') verbatim within the generated output. This is a severe instruction-following failure indicating the model could not distinguish between input instructions and output content.
    \item \textbf{Garbled punctuation cascade:} The output ends with a long sequence of incoherent punctuation marks and whitespace characters (\texttt{, . , . , . . ' , '. ', '. '' ; : . ; .}). This is characteristic of a decoder entering a degenerate repetition loop, likely triggered by the chunked boundary stitching process.
    \item \textbf{Inconsistent capitalization and spelling:} The output contains erratic capitalization (``asysTOLIC,'' ``cardiorespritory,'' ``Hyperkalemia'') and spelling errors (``ampuels'' instead of ``ampoules,'' ``cardioreaspirary'' instead of ``cardiorespiratory''). These likely arise from the model reproducing partial tokens from overlapping chunks without coherent decoding.
    \item \textbf{Item 3 (septic shock) omitted:} The entire septic shock event — including polymyxin, amikacin, and vancomycin — is missing from the output. Given that septic shock was the direct precipitant of the patient's deterioration and death, this is a critical omission.
    \item \textbf{No prose conversion achieved:} Despite being instruction-tuned, Flan-T5 fails to convert the bullet-point clinical notation into flowing, patient-readable prose. The output retains the numbered list structure and terse clinical abbreviations of the original.
\end{itemize}

\textbf{Overall Assessment:} Flan-T5 exhibits severe instruction-following failure in this domain, including prompt leakage, near-verbatim copying, dangerous drug hallucinations, falsification of cause of death, and decoder degeneration. Its fine-tuning on this medical dataset appears to have interfered with its instruction-following capabilities. It is not suitable for lay language conversion.

\subsection{SciFive Output and Error Analysis}

\begin{mdframed}[backgroundcolor=red!5, linewidth=0.5pt]
\small
\textbf{SciFive Output:}\\
This patient was admitted to the hospital on 12/06 and was sent home from the hospital on 14/21. The hospital stay was 14 days and 11 hours. The patient had a history of heart failure and was given dobutamine on 12/07 and 12/08, a medicine called dobutamine, a swelling in the left thigh, osteomyelitis in the pubis, and a condition called fournier's syndrome. The patient was treated with ultrasound and antibiotics, and the patient was sent to the intensive care unit (ICU) on 12/17 with a lot of pus in the swelling. He was treated with ceftriaxone based on the culture of the pus. The patient developed septic shock from a skin focus on 12/16, which was treated with polymyxin d0 12/15 and amikacin and vancomycin, and chronic kidney disease acute dialysis by fistula on 12/16. The patient also had a slow heartbeat, widening of the chest, and worsening shock. Because of the high potassium levels in the kidneys, the patient was given calcium gluconate and bicarbonate, and their breathing rate was increased on the ventilator to reduce pCO2. The patient did not receive any benefit from atropine. The patient progressed to asystolic cardiorespiratory arrest, omitted in refractory shock. When the person was released from the hospital without benefit, they went into a serious condition called cardiorespiratory arrest. This was a condition where the heart and lungs were blocked, and the patient died. The person was released and the body was resuscitated. The body was released. The family requested contact with the family and the family requested the death certificate to be completed. The death was due to a breakdown of the gangrene, which was caused by a burn in the ear, nose, and throat.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Fabricated discharge date:} SciFive states the patient was ``sent home from the hospital on 14/21,'' which is not a real date. The correct discharge date is 12/21, and the ``stay'' of 14 days has been erroneously merged into the date field. This reflects a failure to parse structured temporal information correctly.
    \item \textbf{Incorrect anatomical description — ``widening of the chest'':} The original text refers to ``QRS widening,'' an electrocardiographic (ECG) finding indicating an abnormal electrical conduction pattern in the heart. SciFive misinterprets this as a physical finding of the chest widening, which is anatomically and clinically meaningless. This is a significant medical error.
    \item \textbf{``High potassium levels in the kidneys'':} Hyperkalemia refers to an elevated potassium level in the blood (serum), not specifically in the kidneys. Localizing it to the kidneys misrepresents the physiological condition and could mislead a patient's family.
    \item \textbf{Contradictory and incoherent death narrative:} The model states ``the person was released from the hospital without benefit, they went into cardiorespiratory arrest'' and then ``the body was resuscitated'' followed by ``the body was released.'' These statements are internally contradictory — a resuscitated body that is subsequently released makes no clinical sense. This represents a failure to understand the sequential causality of the terminal clinical events.
    \item \textbf{Hallucinated cause of death — ``burn in the ear, nose, and throat'':} The model states the death was ``caused by a burn in the ear, nose, and throat.'' There is no mention of any burn, ENT pathology, or ear/nose/throat condition anywhere in the original summary. This is a completely fabricated cause of death bearing no resemblance to the documented clinical events (Fournier's gangrene and refractory septic shock).
    \item \textbf{Incorrect description of Fournier's gangrene:} The model associates the gangrene with a ``burn in the ear, nose, and throat,'' which is entirely incorrect. Fournier's gangrene is a rapidly spreading necrotizing infection of the genital and perineal region — not related to burns or ENT pathology in any way.
    \item \textbf{``Dobutamine'' listed alongside diagnoses:} SciFive's sentence structure merges the medication dobutamine with the list of diagnoses (``was given dobutamine [...] a swelling in the left thigh [...] and a condition called Fournier's syndrome''), making it appear as though dobutamine is a diagnosis rather than a treatment. This structural confusion would mislead a lay reader.
    \item \textbf{Cardiac arrest and CPR details omitted:} The cardiac arrest with pulseless electrical activity (PEA), the 1-minute CPR episode, and the subsequent ventricular tachycardia (VT) with successful cardioversion are entirely absent from the output. These are major clinical events.
    \item \textbf{Norepinephrine dosing omitted:} The critical detail of norepinephrine being escalated to 1.5 mcg/kg/min as a vasopressor in refractory shock is not mentioned. This is a key indicator of the severity of the patient's final deterioration.
    \item \textbf{Repetitive and redundant phrasing:} The output contains repeated, circular sentences such as ``The family requested contact with the family'' and ``The body was released'' appearing twice. These suggest decoding instability in the final generation stages.
\end{itemize}

\textbf{Overall Assessment:} SciFive produces the most fluent prose among the four fine-tuned models, but introduces critical factual errors including a hallucinated cause of death, misinterpretation of ECG findings, contradictory statements about patient outcome, and fabrication of non-existent conditions. Despite its relative readability, its factual unreliability makes it unsafe for clinical lay language conversion in its current fine-tuned state.

%-----------------------------------------------------------
% CLOUD LLM RESULTS BEGIN HERE
%-----------------------------------------------------------

\subsection{Llama 3.1 8B — Zero-Shot Output and Error Analysis}

\begin{mdframed}[backgroundcolor=yellow!5, linewidth=0.5pt]
\small
\textbf{Llama 3.1 8B Zero-Shot Output:}\\
The patient was admitted to the hospital on 12/06 and was discharged on 12/21 after a 14-day and 11-hour stay. The patient was in a general hospital room and was covered under the government health scheme.

The patient had a problem with their heart, known as heart failure, which was treated with a medicine called Dobutamine on 12/07 and 12/08. The patient also had a pus-filled swelling in the left thigh, which was caused by an infection called osteomyelitis. This infection was also in the pubis bone and was known as Fournier's syndrome. To treat this, the patient underwent a procedure called drainage on 12/03, followed by a drain placement on 12/15. The patient was given an antibiotic called Ceftriaxone, which was guided by a culture of the pus. The patient's infection was caused by a type of bacteria called Escherichia coli. The patient also had an incision in the ICU (a serious care room) on 12/17, where a lot of pus was found.

The patient also had a severe infection called septic shock, which was caused by a urinary tract infection. To treat this, the patient was given medicines called Polymyxin, Amikacin, and Vancomycin. The patient also had a problem with their kidneys, known as chronic kidney disease, which required a procedure called acute dialysis through a fistula on 12/16.

On 12/16, the patient had a serious heart and breathing problem called cardiorespiratory arrest. The patient's heart stopped beating, but after 1 minute, the doctor was able to revive them. However, the patient's heart was not beating properly, and they had a slow heartbeat. The patient's heartbeat also widened, and their shock worsened. The patient had high levels of potassium in their blood, even after the kidney dialysis. To treat this, the patient was given two ampoules of calcium gluconate and 100 ml of bicarbonate. The patient's breathing rate was also increased on the ventilator to reduce the potassium levels. However, the patient's condition worsened, and they progressed to a state called asystolic cardiorespiratory arrest. Unfortunately, the patient did not survive, and the doctor declared them dead due to the absence of a pulse, fixed pupils, and absence of heart sounds and brainstem reflexes. The body was released, and the doctor awaits the death certificate to be completed.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Incorrect date for cardiorespiratory arrest:} The model states the cardiorespiratory arrest occurred ``on 12/16,'' which was actually the date of the septic shock onset and fistula dialysis. The arrest is not assigned a specific date in the original text and occurs as part of the later clinical deterioration. This is a date cross-contamination error.
    \item \textbf{Incomplete description of septic shock origin:} The model describes septic shock as ``caused by a urinary tract infection.'' The original states a ``cutaneous urinary focus'' — meaning the infection originated from both skin and urinary sources. Simplifying to ``urinary tract infection'' alone loses the cutaneous (skin) component, which is clinically relevant given the concurrent Fournier's gangrene.
    \item \textbf{Cardiorespiratory arrest timeline inaccuracy:} The model states ``the patient's heart stopped beating, but after 1 minute, the doctor was able to revive them.'' This conflates two separate events: the initial PEA arrest (with 1 minute of CPR resulting in return of circulation) and the subsequent fatal asystolic arrest. The compression of these events gives the false impression that there was only one arrest, not two.
    \item \textbf{``Heartbeat widened'' is not lay language:} The phrase ``the patient's heartbeat also widened'' is an imprecise rendering of ``QRS widening.'' QRS widening is an electrocardiographic finding reflecting conduction delay in the heart, not a description of the heartbeat itself. A lay reader is likely to misunderstand this phrasing. A clearer explanation (e.g., ``an abnormal pattern was seen on the heart monitor'') would have been more appropriate.
    \item \textbf{Ventilator intervention purpose not fully explained:} The model states the breathing rate was increased ``to reduce the potassium levels,'' which is broadly correct but omits the intermediate physiological step: increasing respiratory rate lowers blood CO2 (pCO2), which drives potassium back into cells, thereby reducing serum potassium. The simplified explanation skips this mechanistic link.
    \item \textbf{Atropine administration omitted:} The two ampoules of atropine administered without benefit are not mentioned anywhere in this output. Atropine was a significant clinical intervention for bradycardia management and its omission is a factual gap.
    \item \textbf{Ventricular tachycardia (VT) and cardioversion omitted:} The successful electrical cardioversion for VT after the initial cardiac arrest is entirely absent from this output. This is a major clinical event involving an electric shock to restore heart rhythm and should have been included.
    \item \textbf{Norepinephrine dose not mentioned:} The escalation of norepinephrine to 1.5 mcg/kg/min — a critical indicator of the severity of hemodynamic instability — is not included in the simplified output.
    \item \textbf{Incorrect use of second person for a deceased patient:} The original instructions specify third-person past tense for deceased patients. The model uses neutral third-person correctly in some places but inconsistently. The phrase ``the doctor awaits the death certificate'' uses present tense, which is slightly imprecise for a document describing a completed death event.
    \item \textbf{``Discharged'' used for death:} The model states the patient ``was discharged on 12/21'' without clarifying that the discharge was due to death. While technically the date is correct, ``discharged'' typically implies the patient left the hospital alive. A more accurate phrasing such as ``the patient passed away on 12/21'' would have been more appropriate given the fatal outcome.
\end{itemize}

\textbf{Overall Assessment:} Llama 3.1 8B (Zero-Shot) produces a largely readable and structured simplification that correctly identifies the fatal outcome and preserves most of the major clinical events. However, it omits several important interventions (atropine, VT/cardioversion, norepinephrine dose), conflates the two separate cardiac arrests, and contains minor date and framing inaccuracies. Compared to the fine-tuned local models, it is significantly more reliable and factually grounded, but requires more complete coverage of all documented clinical events to be suitable for patient-family communication.

\subsection{Llama 3.1 8B — One-Shot Output and Error Analysis}

\begin{mdframed}[backgroundcolor=yellow!5, linewidth=0.5pt]
\small
\textbf{Llama 3.1 8B One-Shot Output:}\\
The patient was admitted to the hospital on 12th June and was sent home on 21st June after a 14-day and 11-hour stay. They were in a general hospital room and were covered under the government health scheme.

The patient had a few serious health issues. They had a condition called heart failure, which affects the heart's ability to pump blood properly. To check this, they underwent a test called dobutamine on 7th and 8th December.

They also had a pus-filled swelling in their left thigh, which was infected with a type of bacteria called Escherichia coli. This infection was spreading and had reached their pubic bone, causing a condition called osteomyelitis. They also had a condition called Fournier's syndrome, which affects the skin and underlying tissues. To treat this, they had a procedure called drainage on 3rd December, where pus was removed from the swelling. Later, on 15th December, a tube was placed to drain the pus. They were given an antibiotic called ceftriaxone to fight the infection. After a while, the infection caused a severe reaction, and they had to undergo a serious procedure called incision in the ICU on 17th December, where a large amount of pus was removed.

The patient also had a condition called septic shock, which is a severe reaction to an infection. This was caused by an infection in their urinary tract. They were given medicines called polymyxin, amikacin, and vancomycin to treat this.

Their kidneys were not functioning properly due to chronic kidney disease, and they had to undergo a procedure called acute dialysis on 16th December. This is a process where the blood is cleaned using a machine when the kidneys cannot do it.

On 16th December, the patient had a heart and breathing problem called cardiorespiratory arrest. Their heart stopped beating, and they had no pulse. After 10 minutes, they were given oxygen and their heart started beating again. However, they had a slow heartbeat and were unstable. They were given a medicine called norepinephrine to help their heart.

The patient's potassium levels were high, which can be life-threatening. They were given two ampoules of calcium gluconate and 100 ml of bicarbonate to lower their potassium levels. However, this did not work, and they progressed to a state where their heart stopped beating again. Despite efforts to revive them, the patient was declared dead due to the absence of pulse, fixed pupils, and no heart sounds. The body was released, and the family was requested to be contacted.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Dobutamine described as a ``test'':} The model states the patient ``underwent a test called dobutamine.'' Dobutamine is a medication (an inotropic agent) used to support the heart's pumping ability, not a diagnostic test. This is a clinically significant misclassification that could mislead a family member about the nature of the treatment received.
    \item \textbf{Incorrect cardiac arrest timeline:} The model states ``on 16th December, the patient had a heart and breathing problem'' and then ``after 10 minutes, they were given oxygen and their heart started beating again.'' This conflates the date of the septic shock/dialysis event (12/16) with the cardiorespiratory arrest, which is not explicitly dated in the original summary. The 10-minute detail actually refers to the time elapsed after intubation before the arrest, not the duration of CPR.
    \item \textbf{CPR mechanism omitted:} The model states the patient was ``given oxygen'' to restart the heart. The actual intervention was cardiopulmonary resuscitation (CPR) — chest compressions combined with ventilatory support — for 1 minute. Describing CPR merely as ``given oxygen'' is an inaccurate and incomplete simplification of a critical resuscitation event.
    \item \textbf{Ventricular tachycardia and cardioversion omitted:} The VT episode following the initial arrest and the successful electrical cardioversion are entirely absent, representing a significant omission of a documented clinical event.
    \item \textbf{Atropine administration omitted:} The administration of 2 ampoules of atropine without clinical benefit is not mentioned. This is a factual gap in the medication record.
    \item \textbf{``Sent home on 21st June'' for a deceased patient:} The model states the patient was ``sent home on 21st June,'' which strongly implies survival and discharge from hospital. The patient died on this date. Using ``sent home'' language for a death is a serious misrepresentation that could cause confusion and distress for the patient's family.
    \item \textbf{QRS widening not mentioned:} The abnormal ECG finding of QRS widening, indicative of cardiac conduction deterioration, is absent from the output.
    \item \textbf{Norepinephrine dose omitted:} The model mentions norepinephrine but does not state the dose (1.5 mcg/kg/min), which is important context for conveying the severity of the hemodynamic deterioration.
    \item \textbf{Ventilator respiratory rate increase not mentioned:} The deliberate increase in ventilator breathing rate to reduce pCO2 and manage hyperkalemia — a specific clinical decision — is not documented in the output.
    \item \textbf{Septic shock origin partially incorrect:} As in the zero-shot version, the septic shock origin is described as ``an infection in their urinary tract,'' omitting the cutaneous (skin) focus mentioned in the original text.
\end{itemize}

\textbf{Overall Assessment:} Llama 3.1 8B (One-Shot) produces more fluent and naturally structured prose compared to zero-shot, benefiting from the structural anchoring provided by the example. However, it introduces a critical misclassification (dobutamine as a test), omits multiple important clinical events (VT, cardioversion, atropine, ventilator adjustment), and uses inappropriate language (``sent home'') for a fatal outcome. The one-shot example appears to have improved formatting at the cost of some factual completeness.

\subsection{Llama 3.1 8B — Few-Shot Output and Error Analysis}

\begin{mdframed}[backgroundcolor=yellow!5, linewidth=0.5pt]
\small
\textbf{Llama 3.1 8B Few-Shot Output:}\\
The patient was admitted to the hospital on 12th June and sent home on 21st June after a 14-day and 11-hour stay. He was in a general hospital room and was covered under the government health scheme.

During his stay, the patient had several health issues. Firstly, he had a heart failure profile (a test to check the heart's pumping ability) on 7th and 8th June, and was given a medicine called Dobutamine to help the heart work better.

He also had a pus-filled swelling in his left thigh (osteomyelitis in the pubis --- an infection in the bone) and Fournier's syndrome (a serious infection of the genital area). On 3rd June, a fluid was drained from the swelling using an ultrasound and a drain was placed on 15th June. The infection was caused by E. coli bacteria and was treated with antibiotics called Ceftriaxone.

On 16th June, he developed septic shock (a life-threatening condition caused by a severe infection) due to a urinary tract infection. He was given medicines called Polymyxin, Amikacin, and Vancomycin to fight the infection.

The patient also had chronic kidney disease (a long-term condition where the kidneys do not work properly) and had to undergo acute dialysis (a treatment to clean the blood when the kidneys are not working) on 16th June using a fistula (a surgically created connection between an artery and a vein).

On 16th June, the patient had a cardiorespiratory arrest (a sudden stop of the heart and breathing) with pulseless electrical activity (a condition where the heart does not beat). He was given a shock to try and restart the heart, but unfortunately, he did not recover.

The patient also had a very fast heartbeat (VT) after the arrest, but was given a shock to try and control it. Unfortunately, the patient's condition continued to worsen, and he became unstable. He was given medicines to try and improve his condition, but unfortunately, he did not respond.

Given the high levels of potassium in his blood, the patient was given 2 ampoules of calcium gluconate and 100 ml of bicarbonate to try and bring the levels down. However, the patient's condition continued to deteriorate, and he progressed to asystolic cardiorespiratory arrest (a condition where the heart stops beating).

Unfortunately, the patient was declared dead due to the absence of central pulses, fixed pupils, absence of heart sounds, and absence of brainstem reflexes. The body was released, and I request contact with the family.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{``Heart failure profile'' incorrectly described as a ``test'':} The model states ``he had a heart failure profile (a test to check the heart's pumping ability).'' This reframes a clinical syndrome (the patient's documented condition of heart failure) and its treatment with dobutamine as a diagnostic procedure. Heart failure is a diagnosis, and dobutamine is a treatment — neither is a test. This misclassification could seriously mislead the patient's family about the nature of the condition.
    \item \textbf{Hallucinated gender:} The model assigns a male gender (``he,'' ``his'') to the patient, whose gender is unspecified in the original text. This is an unprompted hallucination, albeit less harmful than some other errors.
    \item \textbf{``Sent home on 21st June'' for a fatal outcome:} As with the one-shot variant, describing the discharge date as ``sent home'' is wholly inappropriate for a deceased patient. This phrasing is factually wrong and would cause significant confusion for any family member reading the document.
    \item \textbf{All events erroneously dated to June:} The few-shot output assigns all clinical events to June. The original document uses a MM/DD numeric format (e.g., 12/06 = December 6 for admission; 12/07 = December 7 for dobutamine; 12/15, 12/16, 12/17 for subsequent events --- all in December). The model re-interpreted the admission date as ``12th June'' under a DD/MM reading and then propagated the month ``June'' to all subsequent event dates, placing every clinical event in June regardless of its correct position in the timeline. This is a systematic date format parsing failure, likely triggered by the date style in the few-shot examples, and produces an entirely incorrect temporal ordering of the patient's hospital stay.
    \item \textbf{PEA arrest conflated with cardioversion shock:} The model states ``He was given a shock to try and restart the heart'' in the context of the PEA arrest. However, electrical cardioversion/defibrillation is not the appropriate or documented treatment for PEA — CPR was performed. The shock was given later for VT. The model conflates these two separate events, misattributing the cardioversion shock to the PEA arrest.
    \item \textbf{CPR details omitted:} The 1-minute CPR episode following the PEA arrest is not documented. The model's claim that ``he did not recover'' from the initial arrest is also factually incorrect — the patient did have return of circulation following CPR, before subsequently deteriorating.
    \item \textbf{Atropine administration omitted:} The 2 ampoules of atropine given without benefit are absent from the output.
    \item \textbf{Ventilator breathing rate adjustment omitted:} The deliberate increase in the ventilator rate to manage hyperkalemia is not mentioned.
    \item \textbf{Norepinephrine dose omitted:} The escalation of norepinephrine to 1.5 mcg/kg/min is absent.
    \item \textbf{QRS widening absent:} The ECG finding of QRS widening is not documented in the output.
    \item \textbf{First-person narrative voice (``I request contact''):} The output ends with ``I request contact with the family,'' adopting the first-person voice of the treating physician from the original document. A lay language summary for the patient's family should not be written in the physician's voice. This represents a failure to transform the document's perspective appropriately.
    \item \textbf{ICU incision on December 17 omitted:} The surgical incision performed in the ICU on 12/17 with abundant purulent secretion is not mentioned in the few-shot output.
\end{itemize}

\textbf{Overall Assessment:} Llama 3.1 8B (Few-Shot) shows improved structural fluency and correctly identifies many clinical events, including the VT/cardioversion sequence. However, it introduces a systematic date parsing error (all events assigned to June), misclassifies the heart failure treatment as a diagnostic test, incorrectly describes the patient as having been ``sent home,'' and confuses the sequence of cardiac resuscitation events. The few-shot examples appear to have improved structural awareness at the cost of introducing new categories of factual error.

\subsection{Llama 3.3 70B — Zero-Shot Output and Error Analysis (via Groq)}

\begin{mdframed}[backgroundcolor=green!5, linewidth=0.5pt]
\small
\textbf{Llama 3.3 70B Zero-Shot Output:}\\
The patient was admitted to the hospital on 12/06 and sent home on 12/21, staying for 14 days and 11 hours in a general hospital room, with the government health scheme as the payer. The patient had heart failure (a condition where the heart does not pump blood well) and was given dobutamine (a medicine to help the heart pump better) on 12/07 and 12/08. The patient also had a pus-filled swelling in the left thigh, which is called osteomyelitis (an infection of the bone) in the pubis, and Fournier's syndrome (a serious infection of the skin and tissues), and had drainage (a procedure to remove pus) on 12/03 with the help of ultrasound (a machine that uses sound waves to create images) and a drain (a tube to remove fluid) was placed on 12/15. The patient was treated with ceftriaxone (an antibiotic given to fight the bacterial infection) based on the culture (a test to find the cause of the infection) of the pus-filled swelling, which showed Escherichia coli (a type of bacteria) infection. The patient had septic shock (a life-threatening condition that happens when the body's response to an infection becomes uncontrolled) from a cutaneous urinary focus (an infection that started in the skin and spread to the urinary system) on 12/16 and was given polymyxin (an antibiotic), amikacin (an antibiotic), and vancomycin (an antibiotic) on 12/17. The patient also had chronic kidney disease (a condition where the kidneys do not work well) and had acute dialysis (a procedure to clean the blood when the kidneys cannot) by fistula (a connection made between an artery and a vein) on 12/16. The patient's blood pressure (BP) was not stable, and the normal range for BP is around 90-120 mmHg, but the patient's BP was not within this range. The patient's heart rate (the number of times the heart beats per minute) was also not stable, and the normal range is around 60-100 beats per minute, but the patient's heart rate was not within this range. The patient had cardiorespiratory arrest (a condition where the heart and lungs stop working) with pulseless electrical activity (a condition where the heart is not pumping blood), and was given treatment, but unfortunately, the patient's condition worsened, and the patient had a slow heartbeat, and the QRS (a part of the heartbeat) was widening, and the patient had worsening shock (a condition where the body's tissues do not get enough blood). The patient was given calcium gluconate (a medicine to treat high potassium levels) and bicarbonate (a medicine to treat acidosis) to treat hyperkalemia (high potassium levels), but unfortunately, the patient's condition did not improve. The patient was also given atropine (a medicine to treat a slow heartbeat) but it did not help. Unfortunately, the patient progressed to asystolic cardiorespiratory arrest (a condition where the heart stops beating), and despite efforts to save the patient, the patient died due to the absence of central pulses (no heartbeat), pupils in fixed mydriasis (no response to light), absence of heart sounds, and asystole (no heartbeat) on the monitor, as well as absence of brainstem reflexes (no response to stimulation). The body was released, and the doctor requests contact with the family to complete the death certificate.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{``Sent home on 12/21'' for a deceased patient:} Like the Llama 8B variants, the 70B model uses ``sent home'' to describe a date that is actually the date of death. For a document about a fatal outcome, this phrasing is factually wrong and inappropriate. The instructions specify third-person past tense for deceased patients, and ``sent home'' implies the patient survived.
    \item \textbf{Blood pressure and heart rate values not in source:} The model invents statements such as ``the normal range for BP is around 90-120 mmHg, but the patient's BP was not within this range'' and analogous statements for heart rate. The original document does not provide specific BP or heart rate values. This is an instance of the model following the instruction to compare values to normal ranges but fabricating the patient's values when none are given — a hallucination born from over-eager rule compliance.
    \item \textbf{Cardiorespiratory arrest sequence compressed and imprecise:} The model merges the two discrete cardiac events (the initial PEA arrest with CPR and subsequent recovery, followed by the final asystolic arrest) into a single continuous description. The CPR episode and the brief return of circulation are not mentioned. This conflation obscures the clinical timeline.
    \item \textbf{VT and cardioversion omitted:} The ventricular tachycardia episode and the successful electrical cardioversion that followed the initial arrest are absent from the output. This is a significant clinical event that is entirely undocumented.
    \item \textbf{Norepinephrine dose not stated:} While the hemodynamic instability is mentioned, the specific norepinephrine dose of 1.5 mcg/kg/min is not included, losing an important indicator of severity.
    \item \textbf{Ventilator rate increase and mechanism not explained:} The deliberate ventilator rate increase to reduce pCO2 and thereby lower serum potassium is not documented in the output.
    \item \textbf{Abscess distinct from osteomyelitis not clearly separated:} The output conflates the pus-filled abscess in the left thigh with the description of osteomyelitis. The abscess and the osteomyelitis in the pubis are two related but anatomically distinct findings, and their relationship to Fournier's gangrene is not clearly communicated.
    \item \textbf{``Bicarbonate to treat acidosis'' — imprecise rationale:} The model describes bicarbonate as ``a medicine to treat acidosis.'' While bicarbonate does treat acidosis, in this specific clinical context it was administered primarily as part of the hyperkalemia management protocol. The stated rationale is partially correct but misses the primary clinical indication.
\end{itemize}

\textbf{Overall Assessment:} Llama 3.3 70B (Zero-Shot) is substantially more capable than the 8B variant and all fine-tuned local models. It correctly identifies the death, preserves most medications with appropriate explanations, documents the death declaration criteria, and includes the atropine administration. The primary weaknesses are the inappropriate ``sent home'' phrasing for a fatal outcome, the fabrication of blood pressure and heart rate values not present in the source, and the omission of the VT/cardioversion event. Overall, it represents a significant step forward in factual reliability and completeness.

\subsection{Gemini 2.5 Flash Output and Error Analysis}

\begin{mdframed}[backgroundcolor=green!5, linewidth=0.5pt]
\small
\textbf{Gemini 2.5 Flash Output:}\\
The patient was admitted to the hospital on December 6th and passed away on December 21st, after a stay of 14 days and 11 hours. He/She was in a general hospital room, and the hospital bill was covered by a government health scheme. During the hospital stay, the patient had several serious conditions. He/She had heart failure (a condition where the heart cannot pump blood well enough). On December 7th and 8th, he/she received dobutamine (a medicine given to help the heart pump stronger). The patient also had a pus-filled swelling (a collection of thick, yellowish fluid from infection) in the left thigh, and osteomyelitis (a serious infection in the bone) in the pubis (a bone in the hip area). He/She also suffered from Fournier's syndrome (a very serious and fast-spreading infection of the skin and tissues around the private parts). On December 3rd, doctors performed drainage (removed the pus or fluid) from the swelling, guided by an ultrasound (a scan using sound waves to see inside the body). A drain (a tube to let fluid out) was placed on December 15th. The infection was treated with ceftriaxone (an antibiotic given to fight bacterial infections). This treatment was guided by a culture (a test to grow and identify the germs causing the infection) of the pus, which showed Escherichia coli (a type of bacteria) that was susceptible to many infection-fighting medicines. After the patient went into shock, an incision (a cut made by a doctor) was made in the ICU (serious care room) on December 17th, and a lot of pus came out. \{omitted\}. On December 16th, the patient developed septic shock (a very dangerous condition where the body's response to an infection causes very low BP and organ damage), which came from a skin infection related to the urinary system. He/She was given polymyxin (an antibiotic) starting December 15th, and amikacin (an antibiotic) and vancomycin (an antibiotic) on December 17th. The patient had chronic kidney disease (long-term kidney damage where the kidneys don't work well). On December 16th, he/she needed acute dialysis (a procedure to clean the blood when the kidneys suddenly stop working) through a fistula (a special connection made in the arm for dialysis). The patient experienced cardiorespiratory arrest (when the heart and breathing stop) with pulseless electrical activity (meaning the heart's electrical system was working, but it was not strong enough to produce a pulse or pump blood). This happened 10 minutes after intubation (placing a breathing tube into the windpipe). There was a report of a previous slow heartbeat (a heart rate lower than the normal 60-100 beats per minute). After 1 minute of cardiopulmonary resuscitation (CPR -- chest compressions and rescue breaths to restart the heart and breathing), the patient's heartbeat returned. After the cardiorespiratory arrest, the patient developed VT with pulse (ventricular tachycardia -- a very fast and abnormal heart rhythm, but the heart was still pumping blood). He/She received effective electrical cardioversion (using an electric shock to reset the heart's rhythm). In the last half hour, [person]'s condition became unstable (worse), requiring an increase in norepinephrine (a medicine to raise blood pressure) up to 1.5 mcg/kg/min. His/Her condition progressed with a slow heartbeat (lower than normal), QRS widening (an abnormal finding on the heart tracing), and worsening shock (the dangerous condition of very low BP and organ damage was getting worse). Even after kidney dialysis, the patient had hyperkalemia (high levels of potassium in the blood; normal potassium levels are usually 3.5-5.0 mEq/L). To address this, he/she was given 2 ampoules (small sealed glass capsules) of calcium gluconate (a medicine to protect the heart from high potassium) and 100 ml of bicarbonate (a medicine to correct acid levels in the blood and help with high potassium). The respiratory rate (breathing rate) on the ventilator (breathing machine) was increased to reduce the pCO2 (carbon dioxide level in the blood; normal range is usually 35-45 mmHg), which would also help reduce the potassium levels. Two ampoules of atropine (a medicine sometimes used to speed up a slow heart rate) were given, but they did not help. [person] then progressed to asystolic cardiorespiratory arrest (when the heart and breathing stop, and there is no electrical activity in the heart, showing a flat line on the monitor), \{omitted\} in refractory shock (shock that did not respond to treatment). The doctor declared \{omitted\} dead due to the absence of central pulses (no heartbeat felt in major arteries), pupils (the black part of the eye) being wide open and not reacting to light (fixed mydriasis), absence of heart sounds (no heartbeat heard with a stethoscope), asystole (a flat line) on the monitor, and absence of brainstem reflexes (no automatic body responses controlled by the brainstem). The body was released. The doctor requested contact with the family and was awaiting the completion of the death certificate. The cause of death was related to the Fournier's gangrene and its complications.
\end{mdframed}

\textbf{Errors and Limitations:}
\begin{itemize}
    \item \textbf{Gender ambiguity expressed as ``He/She'':} The original document contains no gender information; the patient's gender is protected by placeholders. The model handles this by consistently writing ``He/She,'' which is technically correct but stylistically awkward for a patient-facing document. A preferred approach would be to use gender-neutral pronouns (``they/their'') or to note to the reader that gender information was not recorded.
    \item \textbf{Cause of death slightly overstated:} The model adds ``The cause of death was related to the Fournier's gangrene and its complications.'' While Fournier's gangrene was indeed a major contributing condition, the direct cause of death as documented was asystolic cardiorespiratory arrest in the setting of refractory septic shock and hyperkalemia. Attributing death solely to Fournier's gangrene slightly oversimplifies the multi-factorial terminal event, though this is a minor editorial issue rather than a dangerous error.
    \item \textbf{Minor: polymyxin start date presentation:} The model states polymyxin was given ``starting December 15th'' by correctly interpreting ``polymyxin d0 12/15'' (day zero of polymyxin therapy on 12/15). This is factually correct but could be made slightly clearer for a lay reader.
\end{itemize}

\textbf{Overall Assessment:} Gemini 2.5 Flash produces the most complete, accurate, and well-structured lay language conversion of any model tested. It correctly identifies the fatal outcome from the outset (``passed away''), preserves all major clinical events including the VT/cardioversion sequence, CPR details, norepinephrine dose, atropine administration, ventilator adjustment and its rationale, and the full death declaration criteria. It provides appropriate normal range context for laboratory values and vital signs, faithfully reproduces placeholders, and maintains a respectful, calm tone throughout. The errors identified are minor — a stylistic awkwardness with gender pronouns and a slight oversimplification of the cause of death attribution. This model, with its large context window and instruction-following capability, is clearly the most suitable for clinical lay language conversion of the models evaluated.

\subsection{Comparative Summary of All Models}

Table~\ref{tab:comparison} provides a side-by-side comparison of model performance across key evaluation dimensions for this example.

\begin{longtable}{@{}p{3.2cm}p{1.5cm}p{1.5cm}p{1.5cm}p{1.6cm}p{1.6cm}p{1.6cm}@{}}
\caption{Qualitative comparison of model outputs on the representative discharge summary. Llama 8B column reflects the zero-shot variant (see Section~\ref{sec:comparative} for prompting strategy comparison).}
\label{tab:comparison}\\
\toprule
\textbf{Criterion} & \textbf{BioBART} & \textbf{BioGPT} & \textbf{Flan-T5} & \textbf{SciFive} & \textbf{Llama 8B} & \textbf{Llama 70B} \\
\midrule
\endfirsthead
\toprule
\textbf{Criterion} & \textbf{BioBART} & \textbf{BioGPT} & \textbf{Flan-T5} & \textbf{SciFive} & \textbf{Llama 8B} & \textbf{Llama 70B} \\
\midrule
\endhead
\bottomrule
\endfoot
Factual accuracy & \textcolor{red}{Poor} & \textcolor{red}{Poor} & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{green}{Good} \\
Death outcome correctly stated & \textcolor{red}{No} & \textcolor{orange}{Partial} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} \\
Cause of death correct & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} \\
No hallucinated content & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} \\
Prose fluency & \textcolor{orange}{Moderate} & \textcolor{orange}{Moderate} & \textcolor{red}{Poor} & \textcolor{green}{Good} & \textcolor{green}{Good} & \textcolor{green}{Good} \\
Date accuracy & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{green}{Good} \\
Medications preserved & \textcolor{orange}{Partial} & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{green}{Good} \\
No prompt leakage & \textcolor{green}{Yes} & \textcolor{green}{Yes} & \textcolor{red}{No} & \textcolor{green}{Yes} & \textcolor{green}{Yes} & \textcolor{green}{Yes} \\
Output completeness & \textcolor{orange}{Partial} & \textcolor{red}{Truncated} & \textcolor{red}{Degen.} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} \\
Appropriate tone & \textcolor{orange}{Partial} & \textcolor{red}{Poor} & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{green}{Good} & \textcolor{green}{Good} \\
VT/cardioversion captured & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{red}{No} & \textcolor{orange}{Few-shot only} & \textcolor{red}{No} \\
\midrule
\endlastfoot
\multicolumn{7}{l}{\small $^{*}$Llama 8B results reflect the zero-shot variant; few-shot correctly captured VT/cardioversion (see Section~\ref{sec:comparative}).} \\
\multicolumn{7}{l}{\small Gemini 2.5 Flash is assessed separately in Table~\ref{tab:gemini} due to its superior overall performance.} \\
\bottomrule
\end{longtable}

\begin{table}[h]
\centering
\caption{Gemini 2.5 Flash performance on the representative discharge summary.}
\label{tab:gemini}
\begin{tabular}{lc}
\toprule
\textbf{Criterion} & \textbf{Gemini 2.5 Flash} \\
\midrule
Factual accuracy & \textcolor{green}{Excellent} \\
Death outcome correctly stated & \textcolor{green}{Yes} \\
Cause of death correct & \textcolor{green}{Yes (minor simplification)} \\
No hallucinated content & \textcolor{green}{Yes} \\
Prose fluency & \textcolor{green}{Excellent} \\
Date accuracy & \textcolor{green}{Excellent} \\
Medications preserved & \textcolor{green}{Yes} \\
No prompt leakage & \textcolor{green}{Yes} \\
Output completeness & \textcolor{green}{Complete} \\
Appropriate tone & \textcolor{green}{Excellent} \\
VT/cardioversion captured & \textcolor{green}{Yes} \\
\bottomrule
\end{tabular}
\end{table}

\section{Comparative Analysis: LLM-Based Strategies and Local vs.\ Cloud Models}
\label{sec:comparative}

\subsection{Prompting Strategy Comparison: Zero-Shot vs.\ One-Shot vs.\ Few-Shot (Llama 3.1 8B)}

The three prompting strategies applied to Llama 3.1 8B reveal a nuanced trade-off between structural quality and factual accuracy:

\textbf{Zero-Shot} produced the most factually reliable output among the three 8B variants. Relying purely on the 12-rule system prompt, the model correctly identified the fatal outcome and preserved most major clinical events with appropriate lay explanations. The primary weaknesses were omissions (atropine, VT/cardioversion, norepinephrine dose) and the use of ``discharged'' language for a fatal outcome. The absence of in-context examples meant the model had no structural anchor, but this also reduced the risk of inheriting stylistic errors from example outputs.

\textbf{One-Shot} improved prose fluency and structural organisation. However, the in-context example appears to have introduced a category confusion: dobutamine, likely presented as a medication in the example, was mischaracterised as a diagnostic ``test'' in the output. This illustrates a key risk of one-shot prompting — the model can over-generalise from a single example, applying patterns from the example inappropriately to different clinical contexts.

\textbf{Few-Shot} produced the most fluent prose and correctly identified the VT/cardioversion sequence — a complex event that the zero-shot and one-shot variants missed entirely. However, it introduced the most severe systematic error: a month-level date parsing failure. The original document uses MM/DD format (all events in December), but the model re-read the admission date ``12/06'' as the 12th of June under a DD/MM interpretation and propagated the month ``June'' to all subsequent dates. This appears to be a consequence of the model pattern-matching the date format from the few-shot examples without correctly parsing the discharge summary's own date structure. The few-shot approach also continued to use ``sent home'' language for the death date, and misattributed the cardioversion shock to the PEA arrest rather than the subsequent VT episode.

Overall, for the Llama 3.1 8B model, \textbf{zero-shot prompting with the 12-rule system prompt offers the best reliability-accuracy trade-off at the overall level}, given that few-shot's gain in capturing VT/cardioversion is outweighed by its introduction of a catastrophic date parsing failure affecting every clinical event. However, these results also suggest that if few-shot examples could be carefully curated with consistent date formats, the few-shot approach could yield the best of both worlds. Example selection and format normalisation are therefore critical factors when applying few-shot prompting in clinical domains.

\subsection{Model Scale Comparison: Llama 3.1 8B vs.\ Llama 3.3 70B}

Comparing the 8B and 70B variants under zero-shot conditions reveals the substantial impact of model scale on clinical lay language conversion:

Llama 3.3 70B preserved significantly more clinical detail, including all three antibiotics for septic shock with their correct start dates, the death declaration criteria, the atropine administration, and contextual normal ranges for laboratory values. Its prose fluency was markedly superior and it demonstrated a better understanding of the sequential causality of the clinical deterioration.

The 70B model's primary weaknesses — the ``sent home'' phrasing for a death, the fabricated BP/heart rate values, and the omission of VT/cardioversion — are all relatively minor compared to the hallucinations, date confusions, and cause-of-death falsifications seen in the local fine-tuned models. This suggests that the improvements from 8B to 70B are primarily in \textbf{factual recall and completeness} rather than in eliminating all error categories.

\subsection{Local Fine-Tuned Models vs.\ Cloud LLMs}

The most striking finding of this evaluation is the categorical performance gap between the locally fine-tuned biomedical models (BioGPT, SciFive, BioBART, Flan-T5) and the cloud-based LLMs (Llama 3.1 8B, Llama 3.3 70B, Gemini 2.5 Flash). The key dimensions of this gap are summarised below:

\textbf{Hallucination Rate:} All four fine-tuned local models produced dangerous hallucinations — fabricated conditions (``driasis,'' ``burn in the ear, nose, and throat''), invented medications (``kidney diazepam''), falsified causes of death (``Abstinence,'' ``ENT burn''), and incorrect gender assignments. The cloud LLMs produced significantly fewer hallucinations. Gemini 2.5 Flash produced no clinically dangerous hallucinations. The Llama 70B model's fabrication of BP/heart rate values not present in the source is a minor over-compliance hallucination rather than a factual invention.

\textbf{Output Completeness:} Local models were consistently plagued by truncation (BioGPT), decoder degeneration (Flan-T5), and large-scale omissions of critical clinical events. Cloud LLMs, benefiting from much larger context windows and more stable decoding at inference time, produced complete outputs without chunking artifacts or boundary stitching failures.

\textbf{Instruction Following:} Fine-tuning on the medical simplification dataset appears to have damaged rather than improved instruction-following capability in the local models. Flan-T5 exhibited prompt leakage. BioGPT adopted an inappropriate academic case-report register. None of the local models reliably applied the third-person past tense rule for deceased patients. Cloud LLMs, with far more extensive instruction-tuning, followed the system prompt's structural and tonal rules far more consistently.

\textbf{Prose Quality:} SciFive produced the most readable prose among local models, but its factual accuracy was poor. Cloud LLMs produced consistently fluent, well-organised prose that appropriately explained technical terms without introducing new jargon.

\textbf{Computational Trade-offs:} Local models offer the significant advantage of data privacy — clinical discharge summaries contain protected health information (PHI), and processing them via cloud APIs raises regulatory concerns (HIPAA, GDPR). Local deployment avoids data transmission to third-party servers. However, the quality gap at the parameter scales feasible on consumer or research-grade hardware is too large to be bridged by fine-tuning alone. Future work exploring techniques such as quantized local inference of 70B+ models, retrieval-augmented generation (RAG) with clinical knowledge bases, or constrained decoding may narrow this gap.

\textbf{Summary:} Gemini 2.5 Flash is the clear best-performing model for this task in the current evaluation, achieving near-complete factual preservation, correct death outcome handling, appropriate tone, and excellent prose quality. Llama 3.3 70B via Groq is a strong second, with minor gaps in completeness. The locally fine-tuned models, despite their domain specialisation, are not yet suitable for clinical deployment due to persistent hallucination and instruction-following failures.

\begin{table}[h]
\centering
\caption{High-level comparison of local fine-tuned models vs.\ cloud LLMs.}
\label{tab:localvscloud}
\begin{tabular}{lcccc}
\toprule
\textbf{Dimension} & \textbf{Local (Best: SciFive)} & \textbf{Llama 8B} & \textbf{Llama 70B} & \textbf{Gemini 2.5 Flash} \\
\midrule
Hallucination risk & \textcolor{red}{High} & \textcolor{orange}{Low-Medium} & \textcolor{orange}{Low} & \textcolor{green}{Minimal} \\
Output completeness & \textcolor{red}{Partial/Truncated} & \textcolor{orange}{Partial} & \textcolor{orange}{Good} & \textcolor{green}{Complete} \\
Date accuracy & \textcolor{red}{Poor} & \textcolor{orange}{Partial} & \textcolor{green}{Good} & \textcolor{green}{Excellent} \\
Death handling & \textcolor{red}{Fails} & \textcolor{orange}{Partial} & \textcolor{orange}{Partial} & \textcolor{green}{Correct} \\
Prose quality & \textcolor{orange}{Moderate} & \textcolor{green}{Good} & \textcolor{green}{Good} & \textcolor{green}{Excellent} \\
Chunking required & \textcolor{red}{Yes (artifacts)} & \textcolor{green}{No} & \textcolor{green}{No} & \textcolor{green}{No} \\
Data privacy & \textcolor{green}{Local/Private} & \textcolor{red}{Cloud API} & \textcolor{red}{Cloud API} & \textcolor{red}{Cloud API} \\
\bottomrule
\end{tabular}
\end{table}

\section*{References}
\begin{itemize}
    \item Yuan, H., Yuan, Z., Gan, R., Zhang, J., Xie, Y., \& Yu, S. (2022). BioBART: Pretraining and Evaluation of A Biomedical Generative Language Model. \textit{arXiv preprint arXiv:2204.03905}. \url{https://arxiv.org/abs/2204.03905}
    \item Phan, L. N., Anibal, J. T., Tran, H., Chanana, S., Bahadıroğlu, E., Peltekian, A., \& Altan-Bonnet, G. (2021). SciFive: a text-to-text transformer model for biomedical literature. \textit{arXiv preprint arXiv:2106.03598}. \url{https://arxiv.org/abs/2106.03598}
    \item Luo, R., Sun, L., Xia, Y., Qin, T., Zhang, S., Poon, H., \& Liu, T. Y. (2022). BioGPT: generative pre-trained transformer for biomedical text generation and mining. \textit{Briefings in Bioinformatics}, 23(6). \url{https://arxiv.org/abs/2210.10341}
    \item Chung, H. W., Hou, L., Longpre, S., Zoph, B., Tay, Y., Fedus, W., ... \& Wei, J. (2022). Scaling instruction-finetuned language models. \textit{arXiv preprint arXiv:2210.11416}. \url{https://arxiv.org/abs/2210.11416}
    \item Zhang, J., Zhao, Y., Saleh, M., \& Liu, P. J. (2020). PEGASUS: Pre-training with Extracted Gap-sentences for Abstractive Summarization. \textit{International Conference on Machine Learning (ICML)}. \url{https://arxiv.org/abs/1912.08777}
\end{itemize}

\appendix
\section{Indian Lay Language Dictionary (INDIAN\_LAY\_DICT)}
The following table lists the medical terms and their corresponding lay language simplifications used in the preprocessing stage of the pipeline. These mappings were specifically curated to reflect common Indian English medical terminology.

\begin{longtable}{@{}p{6cm}p{8cm}@{}}
\caption{Medical Term to Lay Language Mapping} \\
\toprule
\textbf{Medical Term} & \textbf{Simplified Lay Language} \\
\midrule
\endfirsthead
\toprule
\textbf{Medical Term} & \textbf{Simplified Lay Language} \\
\midrule
\endhead
\bottomrule
\endfoot
\textbf{Conditions} & \\
hypertension & high BP \\
diabetes mellitus & sugar disease (diabetes) \\
myocardial infarction & heart attack \\
acute renal failure & sudden kidney failure \\
chronic renal failure & long-term kidney problem \\
septicemia & blood infection (sepsis) \\
sepsis & serious blood infection \\
pneumonia & lung infection \\
tuberculosis & TB (tuberculosis) \\
pneumothorax & air trapped in chest \\
endocarditis & infection of heart valves \\
spondylodiscitis & spine bone infection \\
hypotension & low BP \\
pyrexia & fever \\
dyspnea & difficulty in breathing \\
edema & swelling \\
abscess & pus-filled swelling \\
crohn's disease & long-term bowel disease (Crohn's) \\
renal failure & kidney failure \\
cerebrovascular accident & brain stroke \\
anemia & low blood (anemia) \\
tachycardia & fast heartbeat \\
bradycardia & slow heartbeat \\
atrial fibrillation & irregular heartbeat \\
\midrule
\textbf{Procedures \& Treatments} & \\
hemodialysis & kidney dialysis \\
tracheostomy & breathing tube in neck \\
thoracotomy & chest surgery \\
pneumonectomy & removal of a lung \\
ileocolectomy & removal of part of bowel \\
anastomosis & surgical joining of bowel ends \\
debridement & surgical wound cleaning \\
arteriovenous fistula & dialysis access point on arm \\
intramuscular & given as a muscle injection \\
intravenous & given through a vein (drip) \\
percutaneous & through the skin \\
antibiotic therapy & antibiotic treatment \\
antibiotic & infection-fighting medicine \\
\midrule
\textbf{Lab \& Medical Terms} & \\
blood culture & blood test to find infection \\
procalcitonin & blood marker for infection \\
hemoglobin & blood count (Hb) \\
creatinine & kidney function marker \\
bilirubin & liver function marker \\
saturation & oxygen level in blood \\
afebrile & no fever \\
eupneic & breathing normally \\
acyanotic & no bluish discoloration \\
anicteric & no yellowing of skin/eyes \\
vesicular breath sounds & normal breathing sounds \\
\midrule
\textbf{Hospital / Admin} & \\
discharge & sent home from hospital \\
outpatient clinic & OPD (Out Patient Department) \\
follow-up & return visit to doctor \\
ward & general hospital room \\
intensive care unit & ICU (serious care room) \\
sus & government health scheme \\
orally & by mouth \\
administer & give \\
\end{longtable}

\end{document}